% A workaround to allow relative paths in included subfiles
% that are to be compiled separately
% See https://tex.stackexchange.com/questions/153312/subfiles-inside-a-subfile-using-relative-paths
\providecommand{\main}{..}
\documentclass[\main/thesis.tex]{subfiles}

\begin{document}

% environment for acknowledgements.
\begin{acknowledgements} 
First and foremost, praise be to God who gave me the opportunity to embark
upon this wonderful journey of discovery and learning.
Next, I thank my supervisor, Dr. Abram Hindle, for his guidance, support,
and encouragement throughout my Ph.D. journey. His insightful feedback
helped shape my ideas and refine my work, while his passion for knowledge
and commitment to excellence have been truly inspiring. Beyond academia,
his kindness and understanding of my personal struggles have had a profound
impact on me. His trust in my abilities was what kept me going during the
most challenging moments, and I could not have asked for a better mentor.

My heartfelt thanks also goes to my family for their love, sacrifices, and unwavering support throughout 
my Ph.D. journey. I am also sincerely grateful to my friends for their encouragement, kindness, and constant presence, 
which made this demanding path feel lighter and less lonely.
Lastly, I am also thankful to the members of the EMPLAB for their collaborative spirit, 
readiness to assist, and sharing of valuable feedback. Their support, constructive feedback, and sense of community has greatly enriched my
learning experience and was crucial to my growth.
\end{acknowledgements}

\end{document}

%I would like to express my deepest gratitude to my supervisor, Dr. A, for his constant guidance, support, and encouragement throughout my Ph.D. journey. His thoughtful feedback shaped my ideas and strengthened my work, and his passion for knowledge and dedication to excellence continually motivated me. Beyond his academic mentorship, his kindness, patience, and understanding during some of the most difficult moments of my life meant more to me than I can express. His trust in my abilities gave me the strength to persevere when I doubted myself, and I am profoundly grateful to have had him as my mentor.

%I am deeply thankful to my family, whose unconditional love, sacrifices, and unwavering support carried me through this journey, especially during its most challenging phases. I am also sincerely grateful to my friends for their encouragement, kindness, and constant presence, which made this demanding path feel lighter and less lonely. Finally, I would like to thank the members of my lab for their generosity, collaboration, and willingness to share their time and knowledge. Their support, constructive feedback, and sense of community played an essential role in my growth and progress throughout my Ph.D.