% A workaround to allow relative paths in included subfiles
% that are to be compiled separately
% See https://tex.stackexchange.com/questions/153312/subfiles-inside-a-subfile-using-relative-paths
\providecommand{\main}{..}
\documentclass[\main/thesis.tex]{subfiles}

\begin{document}

% environment for preface
% If you want the preface to not be double-spaced, uncomment the corresponding 
% setting above.
\begin{preface}
Some of the research conducted for this thesis forms part of a research collaboration, led by Professor Dr Abram Hindle at the University of Alberta. The experiments mentioned in chapter 3 were designed by myself and Abdul Ali Bangash. The analysis of results and writeup was conducted by me with assitance from Abdul Ali Bangash. Chapter 3 of this thesis has been published as Bangash, Abdul Ali, Hareem Sahar, Abram Hindle, and Karim Ali. ``On the time-based conclusion stability of cross-project defect prediction models." Empirical Software Engineering 25, No. 6 (2020): 5047-5083. 

In chapter 4, I did the software development and some data collection. In both chapter 4 and chapter 5, I setup the experiments, did the analysis of results and manuscript writing as well. Dr Abram Hindle supervised the work. He was involved in concept formation, methodology design, and also provided feedback for the revision of manuscript. Chapter 4 has been published as Sahar, Hareem, Abdul Ali Bangash, Abram Hindle, and Denilson Barbosa. ``IRJIT: A simple, online, information retrieval approach for just-in-time software defect prediction." Empirical Software Engineering 29, No. 5 (2024): 131. 
Chapter 5 of has not been published yet and is under reviews at Empirical Software Engineering (EMSE) journal.
Chapters 1, 2, and 6 were written entirely by me and have not been published elsewhere. Professor Abram Hindle provided feedback for manuscript revisions.
\end{preface}

\end{document}